---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---

##Loading and preprocessing the data 


**1. Load the data**

- Step 1: Add ggplot2 library and set project working directory

```{r, echo=TRUE}
library(ggplot2)
directory <- c("E:\\Temp\\Coursera\\Reproducible Research\\Extra\\Course Project\\Week 2\\wd")
setwd(directory)
message("... Setting Working Directory ...")
```

  
- Step 2: Download and unzip data file

```{r, echo=TRUE}
fileName <- c("repdata_data_activity.zip")
if(!file.exists(fileName)) {
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", fileName)
        message("... Unzipping archived source data ...")
        unzip(fileName)
} else
        message("... Source file already exists ...")
```

  
- Step 3: Load data into a table

```{r, echo=TRUE}
if(any(dir()=="activity.csv")) {   
        message("... Loading activity data ...")
        message("... This will take a while ;)  ...")
        activity_data <- read.csv("activity.csv", header = TRUE)
        message("... Loading data completed ...")
} else
        message("... File activity.csv doesn't exist ...")
```

  
**2. Process/transform the data into a format suitable for analysis**

```{r, echo=TRUE}
activity_df <- data.frame(steps=activity_data[,1], 
                          datetaken=activity_data[,2], 
                          interval=activity_data[,3])
activity_df2 <- na.omit(activity_df)
```

  
## What is mean total number of steps taken per day?


**1. Calculate the total number of steps taken per day**

```{r, echo=TRUE}
mean_df <- mean(activity_df2$steps)

aggregateSumStepsPerDay <- with(activity_df2, aggregate(activity_df2$steps, 
                                                        by=list(activity_df2$datetaken), 
                                                        sum))
```


**2. Make a histogram of the total number of steps taken each day**

```{r, echo=FALSE, fig.height=4}
names(aggregateSumStepsPerDay) <- c("dates","steps")
g <- ggplot(aggregateSumStepsPerDay, aes(x = steps))
g + geom_histogram(fill="blue", binwidth=1000) +
        xlab("Dates") +
        ylab("Total number of steps ") +
        ggtitle("Total number of steps taken per day")
```


**3. Calculate and report the mean and median of the total number of steps taken per day**

```{r, echo=TRUE}
meanStepsPerDay <- mean(aggregateSumStepsPerDay$steps)
medianStepsPerDay <- median(aggregateSumStepsPerDay$steps)
```


- Mean: **`r meanStepsPerDay`**
- Median: **`r medianStepsPerDay`**


## What is the average daily activity pattern?


**1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**


- Step 1: Get average number of steps taken per interval

```{r, echo=TRUE}
aggregateMeanStepsPerInterval <- with(activity_df2,aggregate(activity_df2$steps, 
                                                             by=list(activity_df2$interval), 
                                                             mean))
```

- Step 2: Plot the data using time series plot (base plot system)  

```{r, echo=FALSE, fig.height=4}
names(aggregateMeanStepsPerInterval) <- c("interval","steps")
plot(aggregateMeanStepsPerInterval,
     type="l",  
     ylab="The average number of steps taken",
     xlab="5-minute interval",
     main="The average daily activity pattern",
     col="Blue")
```


**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**


```{r, echo=TRUE}
maxSteps <- max(aggregateMeanStepsPerInterval$steps)
maxStepsIndex <- which.max(aggregateMeanStepsPerInterval$steps)
maxIntervalSteps <- aggregateMeanStepsPerInterval$interval[maxStepsIndex]
```

- Interval with maximum number of steps: **`r maxIntervalSteps`**


## Imputing missing values

**1. Calculate and report the total number of missing values in the dataset**

```{r, echo=TRUE}
numMissingRecord <- nrow(activity_df) - nrow(activity_df2)
```
- Total number of records with missing values: **`r numMissingRecord`**


**2.Devise a strategy for filling in all of the missing values in the dataset.**

- Strategy: *Use the mean for that 5-minute interval to fill the missing value* 

```{r, echo=TRUE}
newStepsData <- numeric()
for (i in 1:nrow(activity_df)) 
{
        temp <- activity_df[i, ]
        if (is.na(temp$steps)) 
        {
                new_steps <- subset(aggregateMeanStepsPerInterval, interval == temp$interval)$steps
        } else 
        { 
                new_steps <- temp$steps
        }
        newStepsData <- c(newStepsData, new_steps)
}
```

**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**  
```{r, recho=TRUE}
activity_df_new <- data.frame(steps=newStepsData, 
                              datetaken=activity_df$datetaken, 
                              interval=activity_df$interval)
```


**4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.**

- Step 1: Get total average number of steps taken per day from the new dataset

```{r, echo=TRUE}
aggregateSumStepsPerDay2<- with(activity_df_new,aggregate(steps, by=list(datetaken), sum))
```

- Step 2: Plot the filled data using histogram 

```{r, echo=FALSE, fig.height= 4}
names(aggregateSumStepsPerDay2) <- c("dates","steps")
g <- ggplot(aggregateSumStepsPerDay2, aes(x = steps))
g + geom_histogram(fill="red", binwidth=1000) +
        xlab("Dates") +
        ylab("Total number of steps ") +
        ggtitle("Total number of steps taken per day ")
```

- Step 3: Calculate and report the mean and median of the total number of steps taken per day
```{r, echo=TRUE}
meanStepsPerDay2 <-mean(aggregateSumStepsPerDay2$steps)
medianStepsPerDay2 <-median(aggregateSumStepsPerDay2$steps)
```

- Mean: **`r meanStepsPerDay2`**
- Median: **`r medianStepsPerDay2`**


**Do these values differ from the estimates from the first part of the assignment?**  
*No*


**What is the impact of imputing missing data on the estimates of the total daily number of steps?**  
*Not much*


## Are there differences in activity patterns between weekdays and weekends? 

**1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**


- Step 1: Determine whether  given date is a weekday or weekend day

```{r, echo=TRUE}

activity_df4 <- activity_df_new
activity_df4$datetaken <- as.Date(activity_df4$datetaken)

day <- weekdays(activity_df4$datetaken)
level <- vector()

for (i in 1:nrow(activity_df4)) {
        if (day[i] == "Saturday") 
        {
                level[i] <- "Weekend"
        } else if (day[i] == "Sunday") {
                level[i] <- "Weekend"
        } else {
                level[i] <- "Weekday"
        }
}
```


- Step 2: Create a new factor variable in the dataset with two levels 

```{r, echo=TRUE}

activity_df4$level <- level
activity_df4$level <- factor(activity_df4$level)
aggregateSumStepsPerDay3<- with(activity_df4,aggregate(steps, by=list(interval,level), mean))
```


**2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**

```{r, echo=FALSE, fig.height=4}
names(aggregateSumStepsPerDay3) <- c("interval", "daylevel", "steps")
ggplot(aggregateSumStepsPerDay3, aes(interval, steps)) + 
        geom_line(color="blue") + 
        facet_grid(daylevel ~.) +
        xlab("5-minute interval") + 
        ylab("The average number of steps taken")
```